---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Introduction to Modern Mathematical Modeling with R:</h1></center>
<center><h1>A User's Manual to Train Mathematical Consultants</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book by</h3></center>
<center><h3>SSP Shen</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">R Code written by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Compiled and Edited by Joaquin Stawsky</div>
<div align="center">San Diego State University, August 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 3: Linear Models Using Regression and Data

### A Linear Model for the Life Expectancy in France
```{r}
#Life expectancy in France 
dat <- read.csv("~/Desktop/RMathModel/data/LifeExpectancyWorldBank2018clean.csv",
               header=TRUE)

dat[1:3,1:5] # yields first 3 rows and first 5 columns

dim(dat) #displays dimension of data frame

#data from 1960 to 2017, 58 years 
dat[1:5,55:60]
which(dat == "France") # find the index at which "France" exists
dat[76,1:5]

yr <- 1960:1989 # create a sequence of years
le <- dat[76,3:32]
le
ler <- as.numeric(le) #Converts characters values into numeric values
par(mar=c(4.5,4.5,2,1.5))  #Allows user to modify and set graph parameters
plot(yr, ler, ylim=c(70,77),
     xlab="Time", ylab="Life Expectancy",
     main="Life Expectancy in France", 
     cex.main=1.5, cex.lab=1.5, cex.axis=1.5)
lm(ler ~ yr) #Creates linear regression model of ler to yr

abline(lm(ler  ~ yr), col="blue", lwd=2) 
#Adds fitted linear model to the graph

text(1968,76, "Trend = 0.2233 per year", cex=1.5, col="blue")
#Adds text to graph

grid(nx = NULL, ny = NULL)
#Adds grid lines to graph
```

<p>&nbsp;</p>
### Energy Consumption and Heating Degree Data
```{r}
#HDD and kWh: United States Monthly data from Oct. to Sept.
#setwd("~/sshen/mathmodel")

# uncomment next two lines and 'dev.off()' to save plot/figure 
# on to your working directory
#setEPS()
#postscript("figm0302.eps", height=6, width=8) 

par(mar=c(4.2,4.7,3.5,0.8))
hdd <- c(163 ,228 ,343 ,373, 301, 238 ,137 ,84 ,38 ,15,14,34)
kwh <- c(593 ,676 ,1335 ,1149 , 1127, 892 , 538 ,289 , 172, 131,134,134) 
lm(kwh ~ hdd)

plot(hdd, kwh, xlim=c(0,400),ylim=c(0,1500),
     xlab="Climate Index Data: HDD",
     ylab="Energy Index Data: kWh",
     main="Monthly Household Energy Consumption\n vs. Climate Data", 
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5)
abline(lm(kwh ~ hdd), col="red",lwd=2) 
text(140,1300, col="red", cex=1.3, 
     "Linear Model: kWh = 3.317HDD + 53.505") 
text(110,1100, col="red", cex=1.3,
     expression(R^2==0.9663)) 
grid(nx = NULL, ny = NULL)
#dev.off()
```

<p>&nbsp;</p>
### An Example of Linear Model and Data Analysis Using R: A Global Warming Dataset
```{r}
#Read the txt data into R by
dtmean <- read.table("~/Desktop/RMathModel/data/gl_land_nohead.txt",
                   header=F)

#Another way to import the data is to set the R working directory 
#to the same directory where the dataset is located.
#setwd("~/sshen/mathmodel")
#Then import the data by reading the file name
dtmean <- read.table("~/Desktop/RMathModel/data/gl_land_nohead.txt", header=F)
dtmean

dim(dtmean)

yrtime <- dtmean[,1]#Column 1 is time in Year
tmean <- dtmean[,2] #Column 2 is mean temp.
summary(tmean) #Displays summary statistics of data
summary(yrtime)

boxplot(tmean, main="Land Temp. Anomalies")
grid(nx = NULL, ny = NULL)

hist(tmean, 
     main="Histogram: Land Temperature Anomalies",
     xlab="Temp. Anomalies")

tmean
yrtime

reg8014 <- lm(tmean ~ yrtime) #Linear regression
summary(reg8014) #Statistics summary of linear regression model

plot(yrtime, tmean, xlab="Year", ylab="Land Temperature Anomalies", 
     main="Global Annual Mean Land Surface Air Temperature",
     type="o")
abline(reg8014, col="red") #Plot the regression line
text(1930, 0.6, "Linear Temp. Trend = 0.88 °C per century", 
     col="red",cex=1.2)
grid(nx = NULL, ny = NULL)

reg8014$residuals #shows the values of all the residuals

re1 <- reg8014$residuals
#Residual standard error, approximately the SD of residuals
sd(re1)*sqrt((135-1)/(135-2))
```
<p>&nbsp;</p>
### Research Level Exploration for Analyzing the Global Warming Data
```{r}
#Hansen's data analysis
#setwd("~/sshen/mathmodel")
dtmean <- read.table("~/Desktop/RMathModel/data/gl_land_oceanHansen.txt",
                     header=TRUE)
colnames(dtmean)

#a) Statistical summary
summary(dtmean$Anomaly)

#b) Boxplot
boxplot(dtmean$Anomaly,
        main="Boxplot of Hansen's Global Temp. Data 1880-2014", 
        ylab="Temp. [°C]")
grid(nx = NULL, ny = NULL)

#c) Histogram
hist(dtmean$Anomaly, 
     breaks=c(0.1, seq(-0.5, 0.8, 0.1)),
     xlab="Temp. Anomaly [°C]",
     main="Histogram of Hansen's Global Temp.")

#d) Linear regression models
y18802014 <- dtmean$Anomaly
x18802014 <- seq(1880,2014)

#setEPS()
#postscript("figm0306.eps", height=6, width=8) 
par(mar=c(4.2,4.7,2.0,0.8))
plot(x18802014, y18802014, 
     cex.lab=1.5, cex.axis=1.5, cex.main=1.5,
     xlab="Year",ylab="Temperature Anomalies [°C]",
     main="Global Annual Mean Surface Air Temp.: Hansen Data", 
     type="o")
grid(nx = NULL, ny = NULL)

reg18802014 <- lm(y18802014  ~ x18802014)
reg18802014
abline(reg18802014, col="red", lwd=4)
text(1920, 0.70, 
     "1880-2014 Trend = 0.68 °C per century", 
     col="red", cex=1.2) 


#Linear regression for the first 31 years
y18801910 <- dtmean$Anomaly[1:31]
x18801910 <- seq(1880,1910) 
reg18801910 <- lm(y18801910  ~ x18801910) 
reg18801910
segments(x18801910[1], fitted(reg18801910)[1],
         x18801910[31], fitted(reg18801910)[31],
         col="blue", lwd=4) 

#This allows to plot a line segment for the first 31 years.
#But abline(reg18801910,col="blue",lwd=2,xlim=c(1880, 1910))
#plots the regression line for the entire period 1880 - 2014.
text(1920, 0.3, "1880-1910 Trend = -0.60 °C per century", 
     col="blue",cex=1.2)

#Linear regression for the first 71 years
y18801950 <- dtmean$Anomaly[1:71] 
x18801950 <- seq(1880,1950) 
reg18801950 <- lm(y18801950  ~ x18801950)
reg18801950
segments(x18801950[1], fitted(reg18801950)[1],
         x18801950[71], fitted(reg18801950)[71],
         col="green",lwd=5)
text(1920, 0.4, "1880-1950 Trend = 0.37 °C per century", 
     col="green", cex=1.2)

#Linear regression for the first 96 years
y18801975 <- dtmean$Anomaly[1:96] 
x18801975 <- seq(1880,1975) 
reg18801975 <- lm(y18801975  ~  x18801975) 
reg18801975
segments(x18801975[1], fitted(reg18801975)[1],
         x18801975[96], fitted(reg18801975)[96],
         col="black",lwd=2)
text(1920, 0.5, "1880-1975 Trend = 0.36 °C per century", 
     col="black", cex=1.2)

#Linear regression for the first 121 years
y18802000 <- dtmean$Anomaly[1:121] 
x18802000 <- seq(1880,2000) 
reg18802000 <- lm(y18802000  ~ x18802000) 
reg18802000
segments(x18802000[1],fitted(reg18802000)[1],
         x18802000[121],fitted(reg18802000)[121],
         col="purple",lwd=4)
text(1920, 0.6, "1880-2000 Trend = 0.55 °C per century", 
     col="purple", cex=1.2) 
#dev.off()
```