---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Introduction to Modern Mathematical Modeling with R:</h1></center>
<center><h1>A User's Manual to Train Mathematical Consultants</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book by</h3></center>
<center><h3>SSP Shen</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">R Code written by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Compiled and Edited by Joaquin Stawsky</div>
<div align="center">San Diego State University, August 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 8: Stochastic Models

### A Nowhere Differentiable but Everywhere Continuous Model
```{r}
#Wierstrass function
a <- 0.5
b <- 13
m <- 10000
n <- 100
w <- matrix(rep(0,m*n),ncol=n)
x <- seq(-2,2,length=m)
w[,1] <- (a^(+1))*cos(b*pi*x)
for (k in 2:n) w[,k] <- w[,k-1]+(a^(+k))*cos((b^k)*pi*x)
plot(x, w[,n]+1,type="l")
grid(nx = NULL, ny = NULL)

#Riemann's construction
m <- 10000
n <- 10000
wr <- matrix(rep(0,m*n),ncol=n)
x <- seq(-2,2,length=m)
wr[,1] <- sin(x)
for (k in 2:n){
  wr[,k] <- wr[,k-1]+(sin((k^2)*x))/(k^2) 
  }
plot(x, wr[,n],type="l")
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### Brownian Motion
```{r}
#Simple Brownian Motion Code in One Dimension N <- 1000 
N <- 1000
dis <- rnorm(N, 0, 1);
dis <- cumsum(dis);
plot(dis, type = "l",
     main = "Brownian Motion in One Dimension", 
     xlab = "Time", ylab = "Displacement")
grid(nx = NULL, ny = NULL)

#Simple Brownian Motion Code in 2-Dimension N <- 1000; 
N <- 20
xdis <- rnorm(N, 0 ,1);
ydis <- rnorm(N, 0 ,1);
xdis <- cumsum(xdis);
ydis <- cumsum(ydis);
plot(xdis, ydis, type="l", 
     main ="Brownian Motion in 2-Dimension", 
     xlab="x", ylab = "y", 
     col="blue")
grid(nx = NULL, ny = NULL)

#Put arrows on the point locations for the order of particle motion
s <- seq(length(xdis)-1)
arrows(xdis[s], ydis[s], xdis[s+1], ydis[s+1],
       col="red")

#Brownian Motion with Displacement Code in One Dimension N <- 1000 
N <- 1000
dis <- rnorm(N, 0, 1);
at <- rpois(N,1)
for(i in 1:N){ 
  if(at[i] != 0){
  dis[i] <- dis[i]*at[i];
  }
}
dis <- cumsum(dis)
plot(dis, type= "l",
     main= "Brownian Motion in One Dimension with Poisson Arrival Process",
     xlab="Time", ylab="Displacement",
     cex.lab=1.2, cex.axis = 1.2)
grid(nx = NULL, ny = NULL)

#Brownian Motion With Displacement Code in Two Dimension N <- 1000 
N <- 1000
xdis <- rnorm(N, 0,1);
ydis <- rnorm(N, 0,1);
xdis <- cumsum(xdis);
ydis <- cumsum(ydis);
at <- rpois(N,1) 
for(i in N){
  if(at[i] != 0){
    xdis[i] <- xdis[i]*at[i]; ydis[i] <- ydis[i]*at[i] }
}

plot(xdis, ydis, type="l",
     main = "Brownian Motion in Two Dimension with Poisson Arrival Process", 
     xlab="x Displacement" , ylab="y Displacement")
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### Dimension of Koch Curve
```{r}
#TurtleGraphics package 
#install.packages("TurtleGraphics") 
library(grid) 
library("TurtleGraphics")

turtle_init() 
turtle_forward(dist=30)
turtle_backward(dist=10) 
turtle_right(angle=90) 
turtle_forward(dist=10) 
turtle_left(angle=135) 
turtle_forward(dist=14) 
turtle_left(angle=90) 
turtle_forward(dist=14) 
turtle_left(angle=135) 
turtle_forward(dist=10)

#Koch snowflake
koch <- function(s=50, n=6) {
  if (n <= 1)
    turtle_forward(s) 
  else {
    koch(s/3, n-1) 
    turtle_left(60) 
    koch(s/3, n-1) 
    turtle_right(120) 
    koch(s/3, n-1) 
    turtle_left(60) 
    koch(s/3, n-1)
  } 
}

turtle_init(600, 400, "error") 
turtle_do({
  turtle_up() 
  turtle_left(90) 
  turtle_forward(250) 
  turtle_right(180) 
  turtle_down() 
  koch(500, 6)
})
```

<p>&nbsp;</p>
### Use R to Calculate the Fractal Dimension
```{r}
# R method for estimating fractal dimensions for time series:

#install.packages("RandomFields")
#library(RandomFields)
#install.packages("fractaldim")
library(fractaldim)

#Standard normal time series
rf2 <- rnorm(1000)
fd.estimate(rf2, methods="variation", 
            plot.loglog = TRUE, col="blue") 
#This fd.estimate command yields a figure which shows the logN vs log epsilon line
#whose slope is the negative of the Hausdorff dimension we wish to compute.
#D <- 2.02

#Uniformly distributed random time series
rf2 <- rnorm(10000)
fd.estimate(rf2, methods="boxcount", plot.loglog = TRUE, col="blue") 
#D <- 1.89

#Brownian motion time series
set.seed(123)
N <- 1000
T <- 1
delt <- T/N
W <- rep(0,N+1)
for (i in 1:N){
  W[i+1] <- W[i]+rnorm(1)*sqrt(delt)
} 
plot(seq(1,N+1),W,type="l")

for (i in 1:N){
  W[i+1] <- W[i]+rnorm(1)*sqrt(delt)
} 
plot(seq(1,N+1),W,type="l")

fd.estimate(W, methods="boxcount", 
            plot.loglog = TRUE, col="blue") 
#D <- 1.51
```

