---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Introduction to Modern Mathematical Modeling with R:</h1></center>
<center><h1>A User's Manual to Train Mathematical Consultants</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book by</h3></center>
<center><h3>SSP Shen</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">R Code written by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Compiled and Edited by Joaquin Stawsky</div>
<div align="center">San Diego State University, August 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 9: Visualize Mathematical Models By R

### Plot Two Different Time Series on the Same Plot
```{r}
#Plot US temp and prec. times series on the same figure
#plot.new()
par(mar = c(4.2,4.2,2.5,4.1))
Time <- 2001:2010
Tmean <- c(12.06, 11.78,11.81,11.72,12.02,
           12.36,12.03,11.27,11.33,11.66)
Prec <- c(737.11,737.87,774.95,844.55,
          764.03,757.43,741.17,793.50,820.42,796.80)

plot(Time,Tmean,
     type="o", col="red",
     xlab="Year", ylab=expression(paste(T[mean], " [", degree,"C]")),lwd=1.5,
     main="Contiguous U.S. Annual Mean Temperature \nand Total Precipitation")
grid(nx = NULL, ny = NULL)
legend(2000.5,12.42, 
       col=c("red"),lty=1,lwd=2.0, 
       legend=c(expression(T[mean])),bty="n",
       text.font=2,cex=1.0)

#Allows a figure to be overlaid on the first plot
par(new=TRUE)
plot(Time, Prec, 
     type="o",col="blue",
     lwd=1.5,axes=FALSE,xlab="",ylab="") 
legend(2000.5,839, col=c("blue"),lty=1,lwd=2.0,
       legend=c("Precipitation"),bty="n",text.font=2,cex=1.0)

#Suppress the axes and assign the y-axis to side 4
axis(4)
mtext("Precipitation [mm]",side=4,line=2) 
#legend("topleft",col=c("red","blue"),lty=1,legend=c("Tmean","Prec"),cex=0.6)
#Plot two legends at the same time make it difficult to adjust the font size
#because of different scale
```

<p>&nbsp;</p>
### Figure setups: Margins, Fonts, Mathematical Symbols, and More
```{r}
#Margins, math symbol, and figure setups
#plot.new()
par(mar=c(6,4,3,3))
x <- 0.25*(-30:30)
y <- sin(x)
x1 <- x[which(sin(x) >=0)]
y1 <- sin(x1)
x2 <- x[which(sin(x) < 0)]
y2 <- sin(x2)
plot(x1,y1,xaxt="n", xlab="",ylab="",lty=1,type="h",
     lwd=3, tck=-0.02, ylim=c(-1,1), col="red",
     col.lab="purple",cex.axis=1.4) 

lines(x2,y2,xaxt="n", xlab="",ylab="",lty=3,type="h",
      col="blue",lwd=8, tck=-0.02)

axis(1, at=seq(-6,6,2),line=3, cex.axis=1.8)
axis(4, at=seq(-1,1,0.5), lab=c("A", "B", "C", "D","E"),
     cex.axis=2,las=2)

text(0,0.7,font=3,cex=6, "Sine waves", col="darkgreen")
#Itatlic font size 2

mtext(side=2,line=2, 
      expression(y==sin(theta-hat(phi))),cex=1.5, col="blue")
mtext(font=2,
      "Text outside of the figure on side 3",side=3,line=1, cex=1.5)
#Bold font

mtext(font=1, side=1,line=1, 
      expression(paste("Angle in radian: ",
                       theta-phi[0])),cex=1.5, col="red")

par(mar=c(8,6,3,2))
par(mgp=c(2.5,1,0))

plot(1:200/20, rnorm(200),sub="Sub-title: 200 Random Values",
     xlab= "Time", ylab="Random Values", main="Normal Random Values", 
     cex.lab=2.0, cex.axis=1.75, cex.main=2.5, cex.sub=1.5)
grid(nx = NULL, ny = NULL)

par(mgp=c(2,1,0))
plot(sin,xlim=c(10,20))
grid(nx = NULL, ny = NULL)

#A fancy plot of the NOAAGlobalTemp time series
#setwd("~/sshen/mathmodel")
NOAATemp <- read.table('~/Desktop/RMathModel/data/gl_land_oceanHansen.txt', 
                      header = TRUE)
plot.new()
par(mar=c(4,4,3,1))
x <- NOAATemp[,1]
y <- NOAATemp[,2]
z <- rep(-99,length(x))

for (i in 3:length(x)-2){ 
  z[i]<-mean(c(y[i-2],y[i-1],y[i],y[i+1],y[i+2]))
}

n1 <- which(y >= 0)
x1 <- x[n1]
y1 <- y[n1]
n2 <- which(y < 0)
x2 <- x[n2]
y2 <- y[n2]
x3 <- x[2:length(x)-2]
y3 <- z[2:length(x)-2] 

plot(x1,y1,
     type="h",xlim=c(1880,2016),
     lwd=3, tck=0.02, ylim=c(-0.7,0.7),
     ylab="Temperature [°C]", xlab="Time",col="red",
     main="NOAA Global Average Annual Mean Temperature Anomalies")
grid(nx = NULL, ny = NULL)
lines(x2,y2,type="h",lwd=3, tck=-0.02, col="blue") 
lines(x3,y3,lwd=2)
```

<p>&nbsp;</p>
### Plot Two or More Panels on the Same Figure
```{r}
#Plot US temp. and prec. times series on the same figure
par(mfrow=c(2,1))
par(mar=c(0,5,3,1)) #Zero space between (a) and (b)
Time <- 2001:2010
Tmean <- c(12.06, 11.78,11.81,11.72,12.02,
           12.36,12.03,11.27,11.33,11.66) 
Prec <- c(737.11,737.87,774.95,844.55,764.03,
          757.43,741.17,793.50,820.42,796.80)
plot(Time,Tmean,
     type="o",col="red",xaxt="n", xlab="",
     ylab=expression(paste(T[mean]," [",degree,"C]"))) 
grid(nx = NULL, ny = NULL)
text(2006, 12,font=2,"US Annual Mean Temperature", cex=1.5) 
text(2001.5,12.25,"(a)")
#Plot the panel on row 2
par(mar=c(3,5,0,1))

plot(Time, Prec,
     type="o",col="blue",
     xlab="Time",ylab="Prec. [mm]") 
grid(nx = NULL, ny = NULL)
text(2006, 800, font=2, 
     "US Annual Total Precipitation", cex=1.5) 
text(2001.5,840,"(b)")

rm(list=ls()) 
#plot.new()
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), 
       widths=c(3,3),heights=c(2,2))
plot(sin,type="l", xlim=c(0,20)) 
plot(sin,xlim=c(0,10)) 
plot(sin,xlim=c(10,20))
```

<p>&nbsp;</p>
### Basic Principles for an R Contour Plot
```{r}
x <- y <- seq(-1, 1, len=25)
z <- matrix(rnorm(25*25),nrow=25)

par(mar=c(3,3,3,3))
contour(x,y,z, 
        main="Contour Plot of Normal Random Values") 
filled.contour(x,y,z, 
      main="Filled Contour Plot of Normal Random Values") 
filled.contour(x,y,z, color.palette = heat.colors, 
      main="Filled Contour Plot of Normal Random Values \nwith Set Colors")
filled.contour(x,y,z, 
      color.palette = colorRampPalette(c("red", "white", "blue")), 
      main="Filled Contour Plot of Normal Random Values \nwith Specific Colors")
```

<p>&nbsp;</p>
### Plot Contour Color Maps for Random Values on a Map
```{r,fig.width = 8, fig.height = 5.3}
## Figure 9.6
#Plot a 5-by-5 grid global map of standard normal random values 
library(maps)
#plot.new()
#Step 1: Generate a 5-by-5 grid (pole-to-pole, lon 0 to 355) 
Lat <- seq(-90,90,length=37) 
#Must increasing 

Lon <- seq(0,355,length=72) 
#Must increasing

#Generate the random values
mapdat <- matrix(rnorm(72*37),nrow=72)
#The matrix uses lon as row going and lat as column
#Each row includes data from south to north
#Define color
int <- seq(-3,3,length.out=81) 
rgb.palette <- colorRampPalette(c('black','purple','blue','white','green', 
                    'yellow','pink','red','maroon'), interpolate='spline')

#Plot the values on the world map
filled.contour(Lon, Lat, mapdat, color.palette=rgb.palette, 
               levels=int, 
               plot.title=title(xlab="Longitude", ylab="Latitude",
               main="Standard Normal Random Values on a World Map:\n5° Lat-Lon Grid"), 
               plot.axes={ axis(1); axis(2);map('world2', add=TRUE);grid()} )
#filled.contour() is a contour plot on an x-y grid. 
#Background maps are added later in plot.axes={} 
#axis(1) means ticks on the lower side
#axis(2) means ticks on the left side
#Save image with width=800, maintain aspect ratio
```

<p>&nbsp;</p>
### Figure 9.7
```{r,fig.width = 5.5,fig.height = 5}
#Plot a 5-by-5 grid regional map to cover USA and Canada 
Lat3 <- seq(10,70,length=13)
Lon3 <- seq(230,295,length=14) 
mapdat <- matrix(rnorm(13*14),nrow=14) 
int <- seq(-3,3,length.out=81) 
rgb.palette <- colorRampPalette(c('black','purple','blue','white','green',
                    'yellow','pink','red','maroon'), interpolate='spline')

filled.contour(Lon3, Lat3, mapdat, 
               color.palette=rgb.palette, 
               levels=int, plot.title=title(
               main="Standard Normal Random Values on\n a World Map: 5° Lat-Lon Grid",
               xlab="Longitude", ylab="Latitude"),
               plot.axes={axis(1); axis(2);map('world2', add=TRUE);grid()})
```

<p>&nbsp;</p>
### Read .nc File
```{r}
#R plot of NCEP/NCAR Reanalysis PSD monthly temp data .nc file 
#http://www.esrl.noaa.gov/psd/data/gridded/data.ncep.reanalysis.derived.surface.html
rm(list=ls(all=TRUE)) 
#setwd("~/sshen/mathmodel")
# Download netCDF file
# Library
#install.packages("ncdf") 
library(ncdf4)
# 4 dimensions: lon,lat,level,time 
nc <- ncdf4::nc_open("~/Desktop/RMathModel/data/air.mon.mean.nc")
nc
nc$dim$lon$vals # output values 0.0->357.5 

nc$dim$lat$vals # output values 90->-90 

nc$dim$time$vals
#nc$dim$time$units
#nc$dim$level$vals
Lon <- ncvar_get(nc, "lon")
Lat1 <- ncvar_get(nc, "lat")
Time <- ncvar_get(nc, "time")
head(Time)

library(chron)
month.day.year(1297320/24,
            c(month = 1, day = 1, year = 1800)) 
precnc <- ncvar_get(nc, "air")
dim(precnc) #826 months=1948-01 to 2016-10, 68 years 10 months
#plot a 90S-90N precip. along a meridional line at 160E over Pacific 
par(mfrow=c(1,1)) 
plot(seq(90,-90,length=73),precnc[65,,1],
     type="l", xlab="Latitude", 
     ylab="Temperature [°C]",
     main="90S-90N Temperature [°C] Along a Meridional Line at 160E: Jan. 1948",
     lwd=3)
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### Figure 9.9 (a)
```{r,fig.width = 8, fig.height = 5.3}
#Compute and plot climatology and standard deviation Jan 1948-Dec 2015 
library(maps)
climmat <- matrix(0, nrow=144, ncol=73)
sdmat <- matrix(0, nrow=144, ncol=73)

Jmon <- 12*seq(0,67,1) 
for (i in 1:144){
  for (j in 1:73) {
    climmat[i,j] <- mean(precnc[i,j,Jmon])
    sdmat[i,j] <- sd(precnc[i,j,])
  }
}
mapmat <- climmat
#Note that R requires coordinates increasing from south to north -90->90 
#and from west to east from 0->360. We must arrange Lat and Lon this way. 

#Correspondingly, we have to flip the data matrix left to right according to
#the data matrix precnc[i,j,]: 360 (i.e. 180W) lon and from North Pole 
#and South Pole, then lon 178.75W, 176.75W, ..., 0E. This puts Greenwich 
#at the center, China on the right, and USA on the left. However, our map should
#have the Pacific at the center, and USA on the right. Thus, we make a flip .

Lat <- -Lat1

mapmat <- mapmat[,length(mapmat[1,]):1]
#Matrix flip around a column 
#mapmat<- t(apply(t(mapmat),2,rev))
int <- seq(-50,50,length.out=81) 
rgb.palette <- colorRampPalette(c('black','blue','darkgreen',
          'green','white', 'yellow','pink','red','maroon'),
                             interpolate='spline') 

filled.contour(Lon, Lat, mapmat, 
               color.palette=rgb.palette, levels=int,
  plot.title=title(main="NCEP RA 1948-2015 January Climatology [°C]",
               xlab="Longitude",ylab="Latitude"), 
        plot.axes={axis(1); axis(2);map('world2', add=TRUE);grid()}, 
               key.title=title(main="[°C]"))
```

<p>&nbsp;</p>
### Figure 9.9 (b)
```{r,fig.width = 8, fig.height = 5.3}
#Plot standard deviation
#plot.new()
par(mgp=c(2,1,0))
par(mar=c(3,3,2,2))
mapmat <- sdmat[,seq(length(sdmat[1,]),1)]
int <- seq(0,20,length.out=81)
rgb.palette <- colorRampPalette(c('black','blue', 'green','yellow',
         'pink','red','maroon'),interpolate='spline')

filled.contour(Lon, Lat, mapmat, 
               color.palette=rgb.palette, levels=int,
      plot.title = 
      title(main="NCEP 1948-2015 Jan. SAT RA Standard Deviation [°C]",
                    xlab="Longitude", ylab="Latitude"), 
        plot.axes={axis(1); axis(2);map('world2', add=TRUE);grid()}, 
               key.title=title(main="[°C]"))
```

