---
title: "&nbsp;"
output:
  html_document: default
  pdf_document: default
---

<center><h1>Introduction to Modern Mathematical Modeling with R:</h1></center>
<center><h1>A User's Manual to Train Mathematical Consultants</h1></center>
<p>&nbsp;</p>
<center><h3>A Cambridge University Press book by</h3></center>
<center><h3>SSP Shen</h3></center>

<p>&nbsp;</p>
<p>&nbsp;</p>

<div align="center">R Code written by Dr. Samuel Shen, Distinguished Professor</div>
<div align="center">San Diego State University, California, USA</div> 
<div align="center"><https://shen.sdsu.edu></div> 
<div align="center">Email: <sshen@sdsu.edu></div> 
<p>&nbsp;</p>
<div align="center">Compiled and Edited by Joaquin Stawsky</div>
<div align="center">San Diego State University, August 2022</div>

<p>&nbsp;</p>
<p>&nbsp;</p>
<p>&nbsp;</p>
## Chapter 10: Statistical Models and Hypothesis Tests

### Statistical Indices from the Global Temperature From 1880 to 2015
```{r}
#setwd("~/sshen/mathmodel")
dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt") 
dim(dat1)
tmean15 <- dat1[,2] #Take only the second column of this data matrix 

head(tmean15) #The first five values

mean(tmean15)

sd(tmean15) 

var(tmean15)

library(e1071) # R library needed to compute the following parameters

skewness(tmean15)

kurtosis(tmean15)

median(tmean15)

quantile(tmean15,
         probs= c(0.05,0.25, 0.75, 0.95)) 

yrtime15 <- seq(1880,2015)
reg8015 <- lm(tmean15 ~ yrtime15)
reg8015 # Display regression results

# Plot the temperature time series and its trend line
plot(yrtime15, tmean15, 
     xlab="Year",ylab=expression(paste("Temperature [ ", degree, "C]")), 
     main="Global Annual Mean Land and Ocean Surface \nTemperature Anomalies 1880-2015", 
     type="l")
grid(nx = NULL, ny = NULL)
abline(reg8015, col="red")
text(1933, 0.4, 
     expression(paste("Linear Temperature Trend = 0.6678 ", degree, "C per century")), 
     col="red",cex=1.2)
```

<p>&nbsp;</p>
### Histogram of a Set of Data
```{r}
h <- hist(tmean15, #Plot histogram 
      main="Histogram of 1880-2015 Temperature Anomalies",
      xlab="Temperature Anomalies") 

xfit <- seq(min(tmean15),
      max(tmean15), length=30) 
areat <- diff(h$mids[1:2])*length(tmean15) 
#Normalization area 

yfit <- areat*dnorm(xfit, 
        mean=mean(tmean15), sd=sd(tmean15)) 
lines(xfit,yfit,col="blue",lwd=2) 

#Plot the normal fit
plot(density(tmean15), #R estimate density 
     main="R Estimate of Density",xlab="Temperature Anomalies") 

lines(xfit,dnorm(xfit, #Moment estimated normal
      mean=mean(tmean15), sd=sd(tmean15)), col="blue") 
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### Box Plot
```{r}
b <- boxplot(tmean15, ylab="Temperature Anomalies")
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### Scatter Plot
```{r}
#Use setwd("working directory") to work in the right directory
rm(list=ls()) 
#setwd("~/sshen/mathmodel")
par(mgp=c(1.5,0.5,0)) 
ust <- read.csv("~/Desktop/RMathModel/data/USJantemp1951-2016-nohead.csv",
             header=FALSE)

soi <- read.csv("data/soi-data-nohead.csv", 
             header=FALSE) 

soid <- soi[,2] #Take the second column SOI data 
soim <- matrix(soid, ncol=12, byrow=TRUE)
#Make the SOI into a matrix with each month as a column 
soij <- soim[,1] #Take the first column for Jan SOI 
ustj <- ust[,3] #Take the third column: Jan US temp data 
plot(soij, ustj, # Plot the scatter plot
     xlim=c(-4,4), ylim=c(-8,8),
     main="January SOI and the U.S. Temperature", 
     xlab="SOI [dimensionless]",
     ylab=expression(paste("U.S. Temperature [ ", degree, "F]")),
     pch=19, cex.lab=1.3)
grid(nx = NULL, ny = NULL)

soiust <- lm(ustj  ~ soij) 
#Linear regression

abline(soiust, col="red", lwd=3) 
#Linear regression line

#El Nino Years
soijc <- soij[c(1:7,9:32,34:41,43:47,49:65)] 
ustjc <- ustj[c(1:7,9:32,34:41,43:47,49:65)]
```

<p>&nbsp;</p>
### QQ-Plot
```{r}
#setwd("~/sshen/mathmodel")
dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt") 
dim(dat1)
tmean15 <- dat1[,2] 
#qq-plot for standardized anomalies
tstand <- (tmean15 - mean(tmean15))/sd(tmean15)
qqnorm(tstand, 
       ylab=expression(paste("Global Temperature Anomalies [ ", degree, "C]")),
       xlab="Quantile of N(0,1)", 
       xlim=c(-3,3),ylim=c(-3,3)) 
qqline(tstand, col = "red", lwd=2)
grid(nx = NULL, ny = NULL)
```

<p>&nbsp;</p>
### What is a Probability Distribution?
```{r}
#plot.new()
## Figure 10.6
par(mar = c(2.5,4,4,1))
layout(matrix(c(1,2,3), 1, 3, byrow = TRUE),
       widths=c(3,3,3), heights=c(1,1,1)) 

lasvegas <- c(0.58,0.42)
sandiego <- c(0.4,0.6)
seattle <- c(0.16,0.84)

names(lasvegas) <- c("Clear","Cloudy") 
names(sandiego) <- c("Clear","Cloudy") 
names(seattle) <- c("Clear","Cloudy") 

barplot(lasvegas,col=c("skyblue","gray"),ylab="Probability") 
mtext("Las Vegas", side=3,line=1) 

barplot(sandiego,col=c("skyblue","gray"))
mtext("San Diego", side=3,line=1) 

barplot(seattle,col=c("skyblue","gray"))
mtext("Seattle", side=3,line=1) 
mtext("Probability Distribution of Weather",
      cex=1.3,side = 3, line = -1.5, outer = TRUE)
```

<p>&nbsp;</p>
### Figure 10.7
```{r}
# Create data for the area to shade
cord.x <- c(-3,seq(-3,3,0.01),-1)
cord.y <- c(0,dnorm(seq(-3,3,0.01)),0)

# Make a curve
curve(dnorm(x,0,1), xlim=c(-3,3), lwd=3,
      main='PDF of the Standard Normal Distribution',
      xlab='Random Variable x',
      ylab='Probability Density')

# Add the shaded area using many lines
polygon(cord.x,cord.y,col='skyblue') 
polygon(c(-1.5,-1.5, -1.2, -1.2),
        c(0, dnorm(-1.5),dnorm(-1.2), 0.0),col='white')

text(0,0.18, "Area = 1", cex=1.5)
text(-1.65,0.045,"f(x)")
text(-1.35,0.075,"dx")
text(-1.6,0.0095,"x")
text(-0.9,0.0095,"x+dx")

arrows(-2,0.2,-1.35,0.13, length=0.1)
text(-2,0.21,"dA = f(x)dx")
text(0,0.09,expression(paste(integral(f(x)*dx,- infinity,infinity),"=1")))
```

<p>&nbsp;</p>
### Normal Distribution
```{r}
#Normal distribution plot
x <- seq(-8, 8, length=200)

plot(x,dnorm(x, mean=0, sd=1), type="l", lwd=4, col="red",
     ylim = c(0,1),
     xlab="Random Variable x",
     ylab ="Probability Density", 
     main=expression(Normal ~Distribution  ~ N(mu,sigma^2)))
grid(nx = NULL, ny = NULL)
lines(x,dnorm(x, mean=0, sd=2), type="l", lwd=2, col="blue") 
lines(x,dnorm(x, mean=0, sd=0.6), type="l", lwd=2, col="black") 
lines(x,dnorm(x, mean=3, sd=1), type="l", lwd=2, col="purple") 
lines(x,dnorm(x, mean=-4, sd=1), type="l", lwd=2, col="green") 

#ex.cs1 <- expression(plain(sin) * phi, paste("cos", phi)) 
ex.cs1 <- expression(paste(mu, "=0", ~"," ~ sigma, "=1"),
                     paste(mu, "=0", ~"," ~ sigma, "=2"), 
                     paste(mu, "=0", ~"," ~ sigma, "=1/2"),
                     paste(mu, "=3", ~"," ~ sigma, "=1"), 
                     paste(mu, "=-4", ~"," ~ sigma, "=1"))

legend("topleft",legend = ex.cs1, lty=1, 
       col=c('red','blue','black','purple','green'), cex=1, bty="n")

mu <- 0
sig <- 1
intg <- function(x){(1/(sig*sqrt(2*pi)))*exp(-(x-mu)^2/(2*sig^2))} 
integrate(intg,-2,2)

#Or using the R built-in function dnorm to get the same result 
integrate(dnorm,-2,2)

integrate(dnorm,-1.96,1.96)
```

<p>&nbsp;</p>
### Student's t-Distribution
```{r}
#Plot t-distribution by R
x <- seq(-4, 4, length=200)

#plot.new()
plot(x,dt(x, df=3), type="l", lwd=4, col="red", ylim = c(0,0.6),
     xlab="Random Variable t",
     ylab ="Probability Density",
     main="Student t-Distribution T(t,df)")
grid(nx = NULL, ny = NULL)

lines(x,dt(x, df=1), type="l", lwd=2, col="blue") 
lines(x,dt(x, df=2), type="l", lwd=2, col="black") 
lines(x,dt(x, df=6), type="l", lwd=2, col="purple") 
lines(x,dt(x, df=Inf), type="l", lwd=2, col="green") 

#ex.cs1 <- expression(plain(sin) * phi, paste("cos", phi)) 
ex.cs1 <- c("df = 3", "df = 1","df = 2","df = 6",expression(paste("df = ",infinity))) 
legend("topleft",legend = ex.cs1, lty=1,
       col=c('red','blue','black','purple','green'), cex=1, bty="n")

```

<p>&nbsp;</p>
### Probability of a Sample Inside a Confidence Interval
```{r}
#Confidence interval simulation 
mu <- 14 #true mean

sig <- 0.3 #true sd

n <- 50 #sample size 

d <- 1.96*sig/sqrt(n)
lowerlim <- mu-d
upperlim <- mu+d
ksim <- 10000 #number of simulations
k <- 0 # simulation counter
xbar <- 1:ksim
for (i in 1:ksim)
{
  xbar[i] <- mean(rnorm(n, mean = mu, sd = sig))
  if (xbar[i] >= lowerlim & xbar[i] <= upperlim)
    k <- k+1
}

print(c(k,ksim))

#plot the histogram 
#plot.new()
hist(xbar,breaks=51,xlab=expression(paste("Temperature [ ", degree, "C]")),
     main="Histogram of Simulated Sample Mean Temperatures",
     xaxt="n",ylim=c(0,600))

axis(1,pos = -20, at =c(13.92, 14.0, 14.08))
text(14,550,"95% Confidence Interval: (13.92,14.08)",cex=1.2)
#dev.off()
```

<p>&nbsp;</p>
### Confidence Interval of the Sample Mean
```{r}
#Figure of confidence intervals and tail probability
par(mar = c(2.5,3.5,2.0,0.5))
rm(list=ls())
par(mgp=c(1.4,0.5,0)) 
curve(dnorm(x,0,1), xlim=c(-3,3), lwd=3,
      main='Confidence Intervals and Confidence Levels',
      xlab="True Mean as a Random Variable", 
      ylab='Probability Density',xaxt="n", cex.lab=1.3)

polygon(c(-1.96, seq(-1.96,1.96,len=100), 1.96), 
        c(0,dnorm(seq(-1.96,1.96,len=100)),0),col='skyblue')

polygon(c(-1.0,seq(-1.0, 1, length=100), 1),
        c(0, dnorm(seq(-1.0, 1, length=100)), 0.0),col='white')

polygon(c(-3.0,seq(-3.0, -1.96, length=100), -1.96),
        c(0, dnorm(seq(-3.0, -1.96, length=100)), 0.0),col='red')

polygon(c(1.96,seq(1.96, 3.0, length=100), 3.0),
        c(0, dnorm(seq(1.96, 3.0, length=100)), 0.0),col='red')

points(c(-1,1), c(0,0), pch=19, col="blue") 
points(0,0, pch=19) 
points(c(-1.96,1.96),c(0,0),pch=19, col="red")

text(0,0.02, expression(bar(x)), cex=1.0) 
text(-1.50,0.02, "SE", cex=1.0) 
text(-0.60,0.02, "SE", cex=1.0) 
text(1.50,0.02, "SE", cex=1.0)
text(0.60,0.02, "SE", cex=1.0) 
text(0,0.2, "Probability = 0.68") 

arrows(-2.8,0.06,-2.35,0.01, length=0.1) 
text(-2.5,0.09, "Probability")
```

<p>&nbsp;</p>
### Example 10.1
```{r}
#Estimate the mean and error bar for a large sample
#Confidence interval for NOAAGlobalTemp 1880-2015 
#setwd("~/sshen/mathmodel")

dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt") 
dim(dat1)
tmean15 <- dat1[,2]
MeanEst <- mean(tmean15)
sd1 <- sd(tmean15)
StandErr <- sd1/sqrt(length(tmean15))
ErrorMar <- 1.96*StandErr
MeanEst

print(c(MeanEst-ErrorMar, MeanEst+ErrorMar))
```

<p>&nbsp;</p>
### Statistical Inference for x-bar Using a Z-score
```{r}
rm(list=ls())
par(mar = c(2.3,3.0,2.0,0.5))
rm(list = ls())
par(mgp = c(1.0,0.5,0))
curve(dnorm(x,0,1), xlim = c(-3,3), lwd = 3,
      main = 'Z-score, p-value, and Significance Level',
      xlab = "z: Standard Normal Random Variable",
      ylab = 'Probability Density',xaxt = "n",yaxt = "n",
      cex.lab = 1.2,cex.lab = 1.1,cex.axis = 1.1,cex.main = 1.35, ylim = c(-0.1,0.4)) 
lines(c(-3,3),c(0,0))
arrows(-3,-0.1,-2.02,-0.1, lwd = 12,col = 'skyblue', length = 0.2, code = 3)
arrows(3,-0.1,-1.90,-0.1, lwd = 12,col = 'green', length = 0.2, code = 3)
polygon(c(-3.0,seq(-3.0, -2.5, length = 100), -2.5),
        c(0, dnorm(seq(-3.0, -2.5, length = 100)), 0.0),col = 'skyblue')
polygon(c(-1.96,seq(-1.96, 3, length = 100), 3),
        c(0, dnorm(seq(-1.96, 3, length = 100)), 0.0),col = 'lightgreen')
points(-1.96,0, pch = 19, col = "red")
points(-2.5,0,pch = 19, col = "skyblue")
text(-1.4,-0.02, expression(z[0.025]~' = -1.96'), cex = 1.1, col = 'red')
text(-2.40,-0.02, "z-score", cex = 1.1, col = 'skyblue')
arrows(-2.8,0.06,-2.6,0.003, length = 0.1)
lines(c(-1.96,-1.96),c(-0.1, .4),lwd = 1.5, col = 'red')
text(-2.5,0.09, "p-value", cex = 1.3) 
text(1.0,-0.06, expression(H[0] ~'region'), cex = 1.1)
text(-2.5,-0.06, expression(H[1] ~'region'), cex = 1.1)
text(0,0.15, expression(H[0] ~'probability 0.975'), cex = 1.1)
```

<p>&nbsp;</p>
### Example 10.3
```{r}
#Hypothesis test for NOAAGlobalTemp 2006-2015 
#setwd("~/sshen/mathmodel")
dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt")

tm0615 <- dat1[127:136,2]
MeanEst <- mean(tm0615)
MeanEst

sd1 <- sd(tm0615)
sd1

n <- 10
t_score <- (MeanEst -0)/(sd1/sqrt(n))
t_score

1-pt(t_score, df=n-1) #p-value

qt(1-0.025, df=n-1) #critical t-score
```

<p>&nbsp;</p>
### Example 10.4
```{r}
#Hypothesis test for global temp for 1981-1990 and 1991-2000 
#setwd("~/sshen/mathmodel")
dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt") 

tm8190 <- dat1[102:111,2]
tm9100 <- dat1[112:121,2]

barT1 <- mean(tm8190)
barT2 <- mean(tm9100)

S1sd <- sd(tm8190)
S2sd <- sd(tm9100)

n1 <- n2 <- 10

Spool <- sqrt(((n1 - 1)*S1sd^2 + (n2 - 1)*S2sd^2)/(n1 + n2 -2))
t <- (barT2 - barT1)/(Spool*sqrt(1/n1 + 1/n2))

tlow <- qt(0.025, df= n1 + n2 -2)
tup <- qt(0.975, df= n1 + n2 -2)

paste("t-score=", round(t,digits=5),
      "tlow=", round(tlow,digits=5),
      "tup=", round(tup,digits=5))
pvalue <- 1-pt(t, df= n1 + n2 -2)
paste( "p-value=", pvalue)
paste("1981-90 temp=", barT1, "1991-00 temp=",barT2) 

barT2 - barT1
```

<p>&nbsp;</p>
### Statistical Inference of Linear Trend
```{r}
#setwd("~/sshen/mathmodel")
dat1 <- read.table("~/Desktop/RMathModel/data/aravg.ann.land_ocean.90S.90N.v4.0.0.2015.txt") 

tm <- dat1[,2]
x <- 1880:2015
summary(lm(tm ~ x))
```

